---
title: "microbiome-metabolome-curated-data"
output: html_document
date: "2024-05-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(Maaslin2)
library(readr)
library(dplyr)
library(tidyr)

genera <- read_tsv("/Users/una/processed_data/FRANZOSA_IBD_2019/genera.tsv")
metadata <- read_tsv("/Users/una/processed_data/FRANZOSA_IBD_2019/metadata.tsv")

# metadata$Study.Group <- as.character(metadata$Study.Group)
# metadata$Disease.Status <- ifelse(metadata$Study.Group %in% c("UC", "CD"), "Disease", "Control")
```


# Genera-dataset statistics

```{r}
# Convert genera data from wide to long format
genera_long <- gather(genera, key = "Taxon", value = "Abundance", -Sample)

# Calculate genera-dataset statistics
genera.dataset.stats <- genera_long %>%
  group_by(Taxon) %>%
  summarise(
    Taxon.Mean.Abundance = mean(Abundance, na.rm = TRUE),
    Taxon.Perc.of.Non.Zeros = mean(Abundance > 0) * 100
  ) %>%
  ungroup()

# Add averaged statistics
genera.dataset.stats <- genera.dataset.stats %>%
  mutate(
    Averaged.Taxon.Mean.Abundance = Taxon.Mean.Abundance,
    Averaged.Taxon.Perc.of.Non.Zeros = Taxon.Perc.of.Non.Zeros
  )

```

# Discard rare genera (defined here as <25% non-zero values or average abundance <0.1% over all datasets in this analysis)

```{r}
# Filter out genera with averaged percentage of non-zeros less than 25%
genera.dataset.stats <- genera.dataset.stats %>%
  filter(Averaged.Taxon.Perc.of.Non.Zeros >= 25)

# Filter out genera with averaged mean abundance less than 0.001
genera.dataset.stats <- genera.dataset.stats %>%
  filter(Averaged.Taxon.Mean.Abundance >= 0.001)

# Discard genera with percentage of non-zeros less than 10% in individual datasets
genera.dataset.stats <- genera.dataset.stats %>%
  filter(Taxon.Perc.of.Non.Zeros >= 10)

# Discard ambiguous/unidentified genera
genera.dataset.stats <- genera.dataset.stats %>%
  filter(!grepl("g__$", Taxon))

relevant_taxa <- genera.dataset.stats$Taxon
genera_cols <- colnames(genera)
matching_cols <- genera_cols[genera_cols %in% relevant_taxa]
filtered_genera <- genera %>% select(Sample, any_of(matching_cols))

```


# Metabolite-dataset statistics

```{r}
metabolites_map <- read_tsv("/Users/una/processed_data/FRANZOSA_IBD_2019/mtb.map.tsv")
metabolites <- read_tsv("/Users/una/processed_data/FRANZOSA_IBD_2019/mtb.tsv")

# Select only numeric columns for calculation
numeric_metabolites <- metabolites %>% select(where(is.numeric))

# Metabolite-dataset statistics
metabolites.per.dataset <- numeric_metabolites %>%
  summarise_all(list(
    Mean = ~mean(.),
    SD = ~sd(.),
    N = ~n()
  ))

```


# Discard metabolites with no HMDB annotation or metabolites with constant values

```{r}
# Remove rows where HMDB is NA
metabolites_map <- metabolites_map %>% drop_na(HMDB)

# Select only numeric columns for calculation
numeric_metabolites <- metabolites_map %>% select(where(is.numeric))

# Function to check if a metabolite has constant values
is.constant <- apply(numeric_metabolites, MARGIN = 2, function(col) {
  var(col, na.rm = TRUE) == 0})

# Filter out constant metabolites
non_constant_metabolites <- numeric_metabolites[, !is.constant]

# Get the names of the non-constant metabolites
non_constant_metabolite_names <- colnames(non_constant_metabolites)

# Filter the original metabolites dataframe to keep only non-constant metabolites
filtered_metabolites <- metabolites_map %>% select(any_of(non_constant_metabolite_names), everything())

metabolites_cols <- colnames(metabolites)
matching_compounds <- metabolites_cols[metabolites_cols %in% filtered_metabolites$Compound]
filtered_metabolites_data <- metabolites %>% select(Sample, any_of(matching_compounds))

```

# Print statistics

```{r}

common.pairs <- cross_join(genera.dataset.stats,
                           filtered_metabolites) %>%
  group_by(Taxon)

paste(n_distinct(common.pairs), 
      "unique genus-metabolite pairs will be analyzed")

paste("These include", n_distinct(common.pairs$Compound), 
      "metabolites and", n_distinct(common.pairs$Taxon), "genera")

```


# Residuals microbiome

```{r}
row.names(filtered_genera)<-filtered_genera$Sample

metadata <- read_tsv("/Users/una/processed_data/FRANZOSA_IBD_2019/metadata.tsv")
metadata$diagnosis_modified = factor(metadata$Study.Group,
                                              levels = c("Control", "CD", "UC"))

row.names(metadata)<-metadata$Sample

fit_data <- Maaslin2(
    input_data = filtered_genera,
    input_metadata = metadata,
    output = "demo_output",
    fixed_effects = c('diagnosis_modified', 'antibiotic', 'immunosuppressant', 'mesalamine', 'steroids'),
    standardize = FALSE
)

# Extract residuals
residuals_microbiome <- fit_data$residuals

```


# Residuals metabolites

```{r}
row.names(filtered_metabolites_data)<-filtered_metabolites_data$Sample

fit_data_2 <- Maaslin2(
    input_data = filtered_metabolites_data,
    input_metadata = metadata,
    output = "demo_output_2",
    fixed_effects = c('diagnosis_modified', 'antibiotic', 'immunosuppressant', 'mesalamine', 'steroids'),
    standardize = FALSE
)

# Extract residuals
residuals_metabolites <- fit_data_2$residuals

```
